---
title: "Lab1_2024_06_29_Analysis"
output: html_notebook
---

__1) Load Necessary Librarys and Clean Data__
In this section we will import necessary libraries and packages. 
Then, we will load our raw data and select the columns we need for analysis.

```{r}
# Load necessary libraries
library(ggplot2)
library(patchwork)
library(dplyr)

# Load the data from csv file to a data frame
survey_data_org <- read.csv("~/203_Labs_and_Practices/Lab1_Team_project/anes_pilot_2022_csv_20221214.csv")

# Select the columns that will be used in our analysis and store it in a reduced data frame
screened_data <- survey_data_org %>%
  select(caseid, weight, weight_spss, reg, regdiff, turnout22, turnout22ns, pipevote22a, pipevote22b, votehard, vharder_0, vharder_1, vharder_2, vharder_3, vharder_4, vharder_5, vharder_6, vharder_7, vharder_8, vharder_9, vharder_10, vharder_11, vharder_12, vharder_t, waittime, triptime, turnout20, turnout20b, pipevote20, vote20, vote24dt, pid1d, pid1r, pidstr, pidlean, pid_x, imppartyid, votereg, presvote16post, presvote20post, pid3, pid7, ideo5, newsint, iwstartdate, iwstarttime, iwenddate, iwendtime)

# Display the first 5 rows of the reduced data frame to confirm the results
head(screened_data)
```

__2) Filter the valid data and remove those who we shouldn't be considered as 'potential voters'__

Below are the tables and columns we considered in the screening stage of potential voters.

For column 'weight', we remove all records with 'N/A' since we consider it invalid

For column 'reg': Are you registered to vote, or not
            tabulation:  Freq.   Numeric  Label 
                         1,319         1  Yes, registered to vote at my                                         
                            70         2  Yes, registered to vote at a                                          
                           196         3  No, not registered 
Decision: We well discard the columns with value '3' since they are not considered as potential voters.


For column 'turnout22': Did you vote on November 8, 2022?
            tabulation:  Freq.   Numeric  Label 
                           520        ‐1  inapplicable, legitimate skip                            
                           336         1  Definitely voted in person on Nov 8 
                           206         2  Definitely voted in person, before Nov 8 
                           264         3  Definitely voted by mail                            
                           223         4  Definitely did not vote                             
                           36          5  Not completely sure 
Decision: We well use all records including those with value '4' (Definitely did not vote)
The reason why we decided to include all columns of this table is because there may be cases where voters were willing to vote but gave up due to typical difficulties. 

```{r}
# Remove all records with N/As from 'weight'
screened_data <- screened_data %>% filter(!is.na(weight))
# This drops 85 records and 1,500 records are remaining

# Discard all records with value '3' (No, not registered) in column 'reg'
first_screening <- screened_data %>%
  filter(reg != 3)
# Now we have 1,308 observations remaining

```


__3) Adding Republican / Democrat as a column__

Here we wrote codes of identifying whether the voter is a democrat or republican. We decided to use pid_x column
This had the highest response rate by survey takers on which party they most likely fall into. 
We grouped (strong Democrat ~ lean Democrat) to a Democrat denoted as 1 and grouped (lean Republican ~ strong Republican) into another group denoted as 2. 
These will be saved into a new column Political_party_affinity; if independent or skip is used, NA will be placed.

pid_x : Summary of the Party ID
            tabulation:  Freq.   Numeric  Label 
                           371         1  Strong Democrat 
                           183         2  Democrat 
                           141         3  Lean Democrat 
                           273         4  Independent 
                           159         5  Lean Republican 
                           172         6  Republican 
                           284         7  Strong Republican 
                             2         .   


```{r}
# Create a new column for political party affinity
second_screening <- first_screening %>%
  mutate(Political_party_affinity = case_when(
    pid_x %in% c(1, 2, 3) ~ 1,
    pid_x %in% c(5, 6, 7) ~ 2,
    TRUE ~ NA_real_
  ))

# Filter for Republicans and Democrats
filtered_data <- second_screening %>%
  filter(!is.na(Political_party_affinity))
# This leaves us 1,143 observations in final

# Plot the counts of each political party
party_counts <- filtered_data %>%
  count(Political_party_affinity)

Party_Count_Plot <- ggplot(party_counts, aes(x = factor(Political_party_affinity), y = n, fill = factor(Political_party_affinity))) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = n), vjust = -0.5) +
  scale_fill_manual(name = "Political Party", 
                    values = c("1" = "lightblue", "2" = "pink"), 
                    labels = c("1" = "Democrat", "2" = "Republican")) +
  labs(title = "Counts of Political Party Affinity",
       x = "Political Party",
       y = "Number of People") +
  theme_minimal()

print(Party_Count_Plot)
print(party_counts)

```

__4) Now Let's Create Combined Difficulty Analysis for Democrats and Republicans__

Method 1) Use the vharder columns to measure difficulty. 

vharder_1 to vharder_11 contains the results of survey questions asking difficulty on typical subjects.
We will combine these datsets into a single column and measure overall difficulty.

<The original survey questions coming from ANES report>

[vharder0]__Difficulty registering to vote
[vharder1] __ Long wait time at my polling place
[vharder2] __ Concerns about my identification card
[vharder3] __ The voting machine was too complicated to operate
[vharder4] __ Hard to figure out where my polling place was located
[vharder5] __ Takes too long to get to the polling place from where I live
[vharder6] __ Costs too much for transportation to my polling place
[vharder7] __ My work schedule
[vharder8] __ Bad weather
[vharder9] __ Difficult to obtain an absentee ballot
[vharder10] __ Did not have postage to mail back my ballot
[vharder11] __ Other problem [TEXT BOX 60 CHARACTERS]

Coding Explanation)
In the original data set, each individual is supposed to mark '1' for each question if he/she feels that topic difficult and mark '2' if they don't feel difficult. 
Our main idea is to take the sum of all 11 columns and make a combined column named as 'overall_hard'.
Then, if the individual marked all 11 questions as '2' (not difficult), the value would be 22
If the individual marked all 11 questions as '1' (difficult), the value would be 11
So the range is from 11 to 22

However, to make the results easier to understand,
we will convert all '2's into '0's instead.

Then, the value of 'overall_hard' will range from 0 to 11
0 is not difficult at all and  11 would be extremely difficult. 
We will measure the weighted mean of the 2 different subsets of this column and compare how democrats and republicans responded for difficulty in voting.

We are going to make 2 different subsets.

1) Data set of Democrats only
2) Data set of Republicans only

Our final goal is to run a hypothesis test on this two different subsets.


```{r}
# mutate the values of columns 'vharder_1 to 11'. convert all '2's into '0s'
filtered_data <- filtered_data %>%
  mutate(across(vharder_1:vharder_11, ~ ifelse(. == 2, 0, .)))

# Take the sum of all vharder columns and store it in 'overall_hard'
filtered_data <- filtered_data %>%
  mutate(overall_hard = rowSums(select(., vharder_1:vharder_11)))

# create separate datasets for Democrats / Republicans
democrats <- filtered_data %>%
  filter(Political_party_affinity == 1)
# 611 Observations for democrats

republicans <- filtered_data %>%
  filter(Political_party_affinity == 2)
# 532 Observations for republicans

# calculate counts for Democrats / Republicans
combined_overall_hard_counts_dems <- democrats %>%
  count(overall_hard)

combined_overall_hard_counts_reps <- republicans %>%
  count(overall_hard)


# Plot difficulty for Democrats
Dem_Plot <- ggplot(combined_overall_hard_counts_dems, aes(x = factor(overall_hard), y = n)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  geom_text(aes(label = n), vjust = -0.5) +
  labs(title = "Difficulty histogram for Democrats",
       x = "Difficulty",
       y = "Number of people") +
  theme_minimal() +
  theme(legend.position = "none")

# Plot difficulty for Republicans
Rep_Plot <- ggplot(combined_overall_hard_counts_reps, aes(x = factor(overall_hard), y = n)) +
  geom_bar(stat = "identity", fill = "pink") +
  geom_text(aes(label = n), vjust = -0.5) +
  labs(title = "Difficulty histogram for Republicans",
       x = "Difficulty",
       y = "Number of people") +
  theme_minimal() +
  theme(legend.position = "none")

# Plot the results on a single screen
combined_plot <- Dem_Plot + Rep_Plot
print(combined_plot)


```

__5) Now, lets run a hypothesis test with Method 1__

As we see on the distribution plots, we have a highly skewed distribution
Since our dataset is unpaired and
1) ordinal variable (based on survey data with binary scale)
2) IID

We will take the Wilcoxon rank-sum test

First we need to calculate the weigthed mean of each subset. 
We have the weights ranging from 0 to 11 (0 is not difficult at all, 11 is very difficult)
Then this is how we should calculate the weighted mean (expected value)
weighted_mean = sum(count*value) / total_count
For example, the democrat subset's weighted mean would be
(0*361 + 1*112 + 2*51 + 3*23 + 4*10 + 5*3 + 6*2 + 7*2 + 8*1 + 10*2) / 567(total count)

Running multiple experiments, we recognized the wilcoxon rank sum test function calculates the weighted means automatically. 
So, we may simply use that function. 

Below are the codes that we used to be sure that the calculation was correct.

We used 3 different methods and compared the results.
1) run a test with the original data set.
2) calculate the weighted mean manually with mathematic equations
3) run a test with a new column generated as applying the weight to the counts.

The conclusion is that all tests produce the same result.
So we are confident that our test is valid.

```{r}
# Run the wilcoxon rank sum test with the original data set
wilcox_test_result_org <- wilcox.test(democrats$overall_hard, republicans$overall_hard)
print(wilcox_test_result_org)

# Cross check) Check if the wilcox_test result with the original data is same with manual calculation

# Step 1) Calculate the weighted mean manually.
weighted_mean_dem <- sum(combined_overall_hard_counts_dems$overall_hard * combined_overall_hard_counts_dems$n) / sum(combined_overall_hard_counts_dems$n)
weighted_mean_rep <- sum(combined_overall_hard_counts_reps$overall_hard * combined_overall_hard_counts_reps$n) / sum(combined_overall_hard_counts_reps$n)
print(weighted_mean_dem)
print(weighted_mean_rep)
# weighted mean of democrats: 0.7021277, weighted mean of republicans: 0.3515038

# Step 2) Generate a new column with applied weights using 'rep'
dem_scores <- rep(combined_overall_hard_counts_dems$overall_hard, combined_overall_hard_counts_dems$n)
rep_scores <- rep(combined_overall_hard_counts_reps$overall_hard, combined_overall_hard_counts_reps$n)
mean_dem_scores <- mean(dem_scores)
mean_rep_scores <- mean(rep_scores)
print(mean_dem_scores)
print(mean_rep_scores)
# weighted mean of democrats: 0.7021277, weighted mean of republicans: 0.3515038

# Run the Wilcoxon rank sum test with the weight applied columns
wilcox_test_result <- wilcox.test(dem_scores, rep_scores)
print(wilcox_test_result)
# W = 187526, p-value = 3.998e-08


# W = 187526, p-value = 3.998e-08
# The results are identical
```

__6) For the test result above, the reason why the p_value is too low may be a result of the scale.__

To be sure, we scaled down the distribution to 0 to 1 (originally, it is 0 to 11)
If the test produces the exact same value, it means the scale doesn't affect the test result.

```{r}
# Normalize the overall_hard values
combined_overall_hard_counts_dems$normalized_overall_hard <- combined_overall_hard_counts_dems$overall_hard / 11
combined_overall_hard_counts_reps$normalized_overall_hard <- combined_overall_hard_counts_reps$overall_hard / 11

# Use rep function to create vectors of each subset using normalized values
dem_scores_norm <- rep(combined_overall_hard_counts_dems$normalized_overall_hard, combined_overall_hard_counts_dems$n)
rep_scores_norm <- rep(combined_overall_hard_counts_reps$normalized_overall_hard, combined_overall_hard_counts_reps$n)

# Perform the Wilcoxon rank sum test using normalized scores
wilcox_test_result_norm <- wilcox.test(dem_scores_norm, rep_scores_norm)
print(wilcox_test_result_norm)
# W = 187526, p-value = 3.998e-08


# We have found the result is identical.
# So we can reject the null
# The result tells us that the democrats feel more difficulty compared to republicans.
```


__7) In addition, as an alternative method, lets look at votehard which is a single ordinal scale question on voting difficulty__

This is an alternative analysis.
We have conducted this analysis to compare it with our primary analysis
We used 'votehard' column


votehard: How difficult was it for you to vote?
            tabulation:  Freq.   Numeric  Label 
                           417        ‐1  inapplicable, legitimate skip                          
                           1,009         1  Not difficult at all 
                            98         2  A little difficult 
                            48         3  Moderately difficult 
                             7         4  Very difficult 
                             6         5  Extremely difficult 


```{r}

# create separate datasets for Democrats / Republicans
# We will remove all values with -1 since its invalid data (inapplicable, legitimate skip)
democrats_alter <- filtered_data %>%
  filter(Political_party_affinity == 1) %>%
  filter(votehard != -1)
# 521 Observations for democrats

republicans_alter <- filtered_data %>%
  filter(Political_party_affinity == 2) %>%
  filter(votehard != -1)
# 448 Observations for republicans

# calculate counts for Democrats / Republicans
alternative_votehard_dem <- democrats_alter %>%
  count(votehard)

alternative_votehard_rep <- republicans_alter %>%
  count(votehard)


# Plot difficulty for Democrats
Dem_Plot_alter <- ggplot(alternative_votehard_dem, aes(x = factor(votehard), y = n)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  geom_text(aes(label = n), vjust = -0.5) +
  labs(title = "Difficulty histogram for Democrats",
       x = "Difficulty",
       y = "Number of people") +
  theme_minimal() +
  theme(legend.position = "none")

# Plot difficulty for Republicans
Rep_Plot_alter <- ggplot(alternative_votehard_rep, aes(x = factor(votehard), y = n)) +
  geom_bar(stat = "identity", fill = "pink") +
  geom_text(aes(label = n), vjust = -0.5) +
  labs(title = "Difficulty histogram for Republicans",
       x = "Difficulty",
       y = "Number of people") +
  theme_minimal() +
  theme(legend.position = "none")

# Plot the results on a single screen
combined_plot_alter <- Dem_Plot_alter + Rep_Plot_alter
print(combined_plot_alter)

```

__8) Now lets run a wilcoxon rank sum test for the alternative method as well__

The difference is that the weighted mean is a single value and the replicate function creates a vector.

```{r}
# Run the Wilcoxon rank sum test with original data
wilcox_test_result_samp_2 <- wilcox.test(democrats_alter$votehard, republicans_alter$votehard)
print(wilcox_test_result_samp_2)
# W = 128919, p-value = 2.351e-06
```

__9) Compare the result__
The P-value is still extremely low. we can reject the null
(Democrats feel the voting more difficult)

